---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Manching Ho"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---


## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

---

## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.


## Instructions — delete this box



```{r}
my_state <- "CA"        
my_variable <- "B25064_001"    
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```



I obtained 58 rows of data, which matches the total number of counties in California. One county (Alpine County) has missing values for median gross rent.

---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct =moe /estimate* 100)
```



Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```



Sierra County has the largest relative margin of error, with a moe_pct of 59.7%.

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```



Among California counties with valid rent data, 39 are high confidence, 13 are moderate confidence, and 6 are low confidence.

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 15 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```




The average population of counties in the high-reliability group was 989,118, while the average population of counties in the low-reliability group was only 6,771. This pattern shows counties with smaller populations tend to have higher relative error and lower confidence in ACS rent estimates.

---

## What this means for policy



When using data from the ACS to analyze urban policy, we need to pay attention to the margin of error, particularly for small, sparsely populated counties. In California, counties with extremely small populations, such as Sierra County, have very high relative errors, so we cannot rely too heavily on rent estimates from these counties when making policy decisions.

The primary source of this uncertainty is the small sample size of the survey. Because the number of survey respondents is limited, the estimates are subject to greater variation.

Before applying census spatial data in urban planning work, we should check if these estimates are reliable. We cannot take every ACS figure as a hard truth. Recognizing estimation errors prevents us from making poor policy decisions that rely on shaky statistical results.

---



